Dotted Suffix Trees\\A Structure for Approximate Text Indexing

### 2005-09-19 paper/construction.tex
First submission

\begin{enumerate}
\item It does not have the leaf labeled 1 in the original tree. Keeping this leaf would have resulted in the \putstring{.s\$} which does not occur in original string (it is one character too large).
\item For any other leaf $\overline{w\$}$ which occurred at position $p$ in the string, we have a new leaf $\overline{.w\$}$ which occurs at position $p-1$ in the string. Therefore, we adjust the labels.

@@

These conditions are an expression of Lemma~\ref{lemma:suffix-error} and a an extension of the conditions for the root. Both are implemented by Algorithm~\ref{algo:copy-subtree}.

@@

As discussed above, suffix trees are implemented such that each node is actually just a pair of indices into the original string (examplified in Figure~\ref{fig:mississippi-nodes}). To merge a child with its parent, we just need to replace both by a node whose starting index is the starting index of the original child minus the length of the parent edge. This process is illustrated in Figure~\ref{fig:merge}.

@@

The code for the subtree copy is Algorithm~\ref{algo:copy-subtree-with-dot-link} which differs from Algorithm~\ref{algo:copy-subtree} in that dot link are also followed while copying the tree.

The final procedure for constructing k-error trees is Algorithm~\ref{algo:construct-k}.

>>>>

\begin{enumerate}
\item It does not have the leaf labeled 1 in the original tree. Keeping this leaf would have resulted in a leaf labeled \putstring{.s\$} which does not occur in original string (it is one character too long).
\item For any other leaf $\overline{w\$}$ occuring at position $p$ in the string, we have a new leaf $\overline{.w\$}$ which occurs at position $p-1$ in the string. Therefore, we adjust the labels.

@@

These conditions are an expression of Lemma~\ref{lemma:suffix-error} and an extension of the conditions for the root. Both are implemented by Algorithm~\ref{algo:copy-subtree}.

@@

As discussed above, suffix trees are implemented such that each node is actually just a pair of indices into the original string (exemplified in Figure~\ref{fig:mississippi-nodes}). To merge a child with its parent, we just need to replace both by a node whose starting index is the starting index of the original child minus the length of the parent edge. This process is illustrated in Figure~\ref{fig:merge}.

@@

The code for the subtree copy is Algorithm~\ref{algo:copy-subtree-with-dot-link} which differs from Algorithm~\ref{algo:copy-subtree} in that dot link are also followed while copying the tree. The final procedure for constructing k-error trees is Algorithm~\ref{algo:construct-k}.

### 2005-09-19 paper/experimental.tex
First submission

To experimentally verify the average case prediction, we show in Figures~\ref{fig:ratios} the ratios between the k-error and the (k+1)-error dotted trees, for english text (the other datasets show similiar behaviour). We can easily see that the experimental values do resemble a logarithm as predicted.

@@

For searching, the full texts were first indexed and then online searches were performed on top of the structure. To generate input patterns, randomly chosen snippets of the text were taken. These patterns were then altered by substituting a character at random. To generate mismatches, one character too many was altered~\footnote{Since the characters used were not present in the original text, there is no risk of generating a string which \emph{does appear} in the text.}. The search algorithm performed an early exit, without reporting of occurrences (ie, it only reported whether the string exists in the text). Therefore, the number of occurrences had no influence on the search time.

>>>>

To experimentally verify the average case prediction, we show in Figures~\ref{fig:ratios} the ratios between the k-error and the (k+1)-error dotted trees, for English text (the other datasets show similar behaviour). We can easily see that the experimental values do resemble a logarithm as predicted.

@@

For searching, the texts were first indexed and then online searches were performed on top of the structure. To generate input patterns, randomly chosen snippets of the text were taken. These patterns were then altered by substituting a few characters at random. To generate mismatches, one character too many was altered\,\footnote{Since the characters used were not present in the original text, there is no risk of generating a string which \emph{does appear} in the text.}. The search algorithm performed an early exit, without reporting of occurrences (ie, it only reported whether the string exists in the text). Therefore, the number of occurrences had no influence on the search time.

### 2005-09-19 paper/search.tex
First submission

To search a string with at maximum k errors in a k-error dotted tree, we descend the tree. Whenever we find a node we follow both the dot link and the edge which matches the current position of the string we are searching. If we find a mismatch in the middle of an edge, we continue, decreasing the amount of errors permitted.
When presenting the algorithm, we take the liberty of using $s+j$ where $s$ is a string and $j$ an integer to mean the suffix of $s$ starting at position $j$ (as happens in the C~Programming Language).

@@

This algorithm runs in time $\mathcal{O}(m^{k+1})$: The recursive call at line~\ref{algoline:search:followDot} corresponds to following a dot link. There are at most ${m \choose k}=\mathcal{O}(m^k)$ different paths which must be followed. Following one path in the tree takes $\mathcal{O}(m)$ time. Therefore, the total time is $\mathcal{O}(m^{k+1})$ without the reporting of occurrences (which can be done in time linear to the number of occurrences).

>>>>

To search a string with at maximum k errors in a k-error dotted tree, we descend the tree. Whenever we find a node, we follow both the dot link and the edge which matches the current position of the string we are searching for. If we find a mismatch in the middle of an edge, we continue, decreasing the amount of errors permitted. When presenting the algorithm, we take the liberty of using $s+j$ where $s$ is a string and $j$ an integer to mean the suffix of $s$ starting at position $j$ (as happens in the C~Programming Language).

@@

This algorithm runs in time $\mathcal{O}(m^{k+1})$: There are at most $\sum_{i=0}{k}{m \choose i}=\mathcal{O}(m^k)$ different paths which must be followed. Following one path in the tree takes $\mathcal{O}(m)$ time. Therefore, the total time is $\mathcal{O}(m^{k+1})$ without the reporting of occurrences (which can be done in time linear to the number of occurrences).

### 2005-09-19 paper/space.tex
First submission

\begin{definition}[Maximum Path Lenght]
$l$ is the length of the deepest branching node in the tree

@@

We now show $N_k = \mathcal{O}(nl^k)$ using induction. It is a known fact that $N_0 = \mathcal{O}(n)$. The algorithm for turning a k-error into a (k+1)-error dotted tree, can be looked at the following way~\footnote{We are here considering that the second \textit{foreach} loop of Algorithm~\ref{algo:construct-k} runs in a specific order which is in fact difficult to code for. However, as an analysis tool, it is a valid assumption.}: First it makes an almost exact copy of the k-error dotted tree and as the error tree at the root and it clears all the other error trees. Then it proceeds in stages, making a (possibly incomplete) copy of this tree spread amongst the nodes at suffix-depth 1. It proceeds to the other nodes in increasing suffix-depths up to $l$. At each such step, the number of nodes is increased by a maximum of $N_k$. Therefore, we start with $N_k$ nodes, make an almost full copy, and copy that at most $l$ times, $N_{k+1}=\mathcal{O}(N_k(l+1))$. Assuming $N_k=\mathcal{O}(nl^k)$ by induction we conclude $N_{k+1}=\mathcal{O}(nl^{k+1})$.

>>>>

\begin{definition}[Maximum Path Lenght]
$l$ is the string depth of the deepest branching node in the tree:

@@

We now show $N_k = \mathcal{O}(nl^k)$ using induction. It is a known fact that $N_0 = \mathcal{O}(n)$. The algorithm for turning a k-error into a (k+1)-error dotted tree, can be looked at the following way\,\footnote{We are here considering that the second \textit{foreach} loop of Algorithm~\ref{algo:construct-k} runs in a specific order which is in fact difficult to code for. However, as an analysis tool, it is a valid assumption.}: First it makes an almost exact copy of the k-error dotted tree and as the error tree at the root and it clears all the other error trees. Then it proceeds in stages, making a (possibly incomplete) copy of this tree spread amongst the nodes at suffix-depth 1. It proceeds to the other nodes in increasing suffix-depths up to $l$. At each such step, the number of nodes is increased by a maximum of $N_k$. Therefore, we start with $N_k$ nodes, make an almost full copy, and copy that at most $l$ times, $N_{k+1}=\mathcal{O}(N_k(l+1))$. Assuming $N_k=\mathcal{O}(nl^k)$ by induction we conclude $N_{k+1}=\mathcal{O}(nl^{k+1})$.

### 2005-09-19 paper/structure.tex
First submission

\begin{definition}[Character, string]
Given a set $\Sigma$, we say that $s$ is a \emph{string over $\Sigma$} if $s$ is a (possibly empty) sequence of elements of $\Sigma$. The elements of the set $\Sigma$ are also called \emph{characters}. The length of the string, denoted by $|s|$ is the number of elements it contains. We shall write $s_i$ for the $i$th element of the string. For the empty string, we write $\epsilon$.

The set of all strings is denoted by $\Sigma^*$ and $\Sigma^+=\Sigma^*-\{\epsilon\}$.
\end{definition}

@@

\begin{definition}
$wx$ will denote the string formed by all the characters of $w$ followed by all the characters of $x$ (the usual concatenation operation). $aw$ ($wa$) shall denote the string formed by the character $a$ before (after) the characters of $w$.

@@

\begin{definition}[Aproximate Match]
We say that the string $s$ matches the string $t$ at position $p$ with $k$ errors if we can make $k$ substitutions in $s$ to obtain $s'$ is a substring of $t$ at position $p$.
\end{definition}

@@

It is important to understand that suffix trees are normally implemented so that each node consists of only two indices into the original string which are the start and end position of its incoming edge. Figure~\ref{fig:mississippi-nodes} illustrates this.

@@

\begin{lemma}[Position set at the suffix node]\label{lemma:suffix-error}
Given two nodes $\overline{aw}$ and $\overline{w}$ (connected by a suffix link, labeled $a$), if one takes position set of $\overline{w}$, subtracts one from each element, resulting a superset of the position set of $\overline{w}$. The items shared by both sets are those positions of the string which contain an $a$.
\end{lemma}

@@

\begin{definition}[Error Tree]
For any node $w$ its error tree is the Patricia tree formed by taking its position set, adding one to each element and forming the Patricia tree of the suffixes of the suffixes starting at those positions. If the position set includes the end of the string, that element is removed.

>>>>

\begin{definition}[Character, string]
Given a set $\Sigma$, we say that $s$ is a \emph{string over $\Sigma$} if $s$ is a (possibly empty) sequence of elements of $\Sigma$. The elements of the set $\Sigma$ are also called \emph{characters}. The length of the string, denoted by $|s|$ is the number of elements it contains. We shall write $s_i$ for the $i$th element of the string. For the empty string, we write $\varepsilon$.

The set of all strings is denoted by $\Sigma^*$ and $\Sigma^+=\Sigma^*-\{\varepsilon\}$.
\end{definition}

@@

\begin{definition}[Concatenation]
$wx$ will denote the string formed by all the characters of $w$ followed by all the characters of $x$ (the usual concatenation operation). $aw$ ($wa$) shall denote the string formed by the character $a$ before (after) the characters of $w$.

@@

\begin{definition}[Aproximate Match]
We say that the string $s$ matches the string $t$ at position $p$ with $k$ errors if we can make $k$ substitutions in $s$ to obtain $s'$ which is a substring of $t$ at position $p$.
\end{definition}

@@

Suffix trees are normally implemented so that each node consists of only two indices into the original string which are the start and end position of its incoming edge. Figure~\ref{fig:mississippi-nodes} illustrates this.

@@

\begin{lemma}[Position set at the suffix node]\label{lemma:suffix-error}
Given two nodes $\overline{aw}$ and $\overline{w}$ (connected by a suffix link, labeled $a$), if one takes position set of $\overline{w}$, subtracts one from each element, one obtains a superset of the position set of $\overline{w}$. The items shared by both sets are those positions of the string which contain an $a$.
\end{lemma}

@@

\begin{definition}[Error Tree]
For any node $w$, its error tree is the Patricia tree formed by taking its position set, adding one to each element and forming the Patricia tree of the suffixes of the suffixes starting at those positions. If the position set includes the end of the string, that element is removed.
